\renewcommand{\bibname}{References}
\addcontentsline{toc}{chapter}{References}

\begin{thebibliography}{99}

\bibitem{kamble2025}
Kamble, V.\ B., Pisal, K., Vaidya, P., and Gaikwad, S.,
``Enhancing UPI Fraud Detection: A Machine Learning Approach Using Stacked Generalization,''
\textit{International Journal of Multidisciplinary on Science and Management},
vol.\ 2, no.\ 1, pp.\ 69--83, 2025.

\bibitem{khedekar2025}
Khedekar, A.\ and Manerkar, G.,
``Machine Learning-Driven Detection of UPI Fraudulent Activities,''
\textit{International Research Journal of Engineering and Technology (IRJET)},
vol.\ 12, no.\ 4, 2025.

\bibitem{malode2025}
Malode, S., Wankhede, A., Gujarkar, P., Borkar, I., Kakde, G., and Naik, S.,
``UPI Fraud Detection by Using Machine Learning,''
\textit{International Journal for Research Publication and Seminar},
vol.\ 16, no.\ 1, pp.\ 921--924, 2025.

\bibitem{hu2022}
Hu, X., Chen, H., Liu, S., Li, X., Zhang, S., Wang, Y., and Xue, X.,
``Cost-Sensitive GNN-Based Imbalanced Learning for Mobile Social Network Fraud Detection,''
\textit{IEEE Transactions on Computational Social Systems},
pp.\ 1--12, 2022.

\bibitem{chen2016}
Chen, T.\ and Guestrin, C.,
``XGBoost: A Scalable Tree Boosting System,''
in \textit{Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (KDD '16)},
San Francisco, CA, USA, pp.\ 785--794, 2016.

\bibitem{breiman2001}
Breiman, L.,
``Random Forests,''
\textit{Machine Learning},
vol.\ 45, no.\ 1, pp.\ 5--32, 2001.

\bibitem{chawla2002}
Chawla, N.\ V., Bowyer, K.\ W., Hall, L.\ O., and Kegelmeyer, W.\ P.,
``SMOTE: Synthetic Minority Over-sampling Technique,''
\textit{Journal of Artificial Intelligence Research},
vol.\ 16, pp.\ 321--357, 2002.

\bibitem{kaggle2024}
Choudhary, Y.,
``UPI Transaction 2024: Advanced Analysis Dataset,''
Kaggle Dataset, 2024. [Online]. Available:
\url{https://www.kaggle.com/code/yogeshwarchoudhary/upi-transaction-2024-advanced-analysis}

\bibitem{npci2024}
National Payments Corporation of India (NPCI),
``UPI Product Statistics,''
[Online]. Available: \url{https://www.npci.org.in/what-we-do/upi/product-statistics},
Accessed: April 2025.

\bibitem{pagerank1998}
Page, L., Brin, S., Motwani, R., and Winograd, T.,
``The PageRank Citation Ranking: Bringing Order to the Web,''
Technical Report, Stanford InfoLab, 1998.

\bibitem{networkx2008}
Hagberg, A., Swart, P., and Schult, D.,
``Exploring Network Structure, Dynamics, and Function using NetworkX,''
in \textit{Proceedings of the 7th Python in Science Conference (SciPy 2008)},
Pasadena, CA, USA, pp.\ 11--15, 2008.

\bibitem{fastapi2019}
Ram\'irez, S.,
``FastAPI: Modern, Fast (High-Performance), Web Framework for Building APIs with Python 3.7+,''
[Online]. Available: \url{https://fastapi.tiangolo.com}, 2019.

\bibitem{streamlit2019}
Treuille, A., Teixeira, T., and Kelly, A.,
``Streamlit: The Fastest Way to Build and Share Data Apps,''
[Online]. Available: \url{https://streamlit.io}, 2019.

\bibitem{bolton2002}
Bolton, R.\ J.\ and Hand, D.\ J.,
``Statistical Fraud Detection: A Review,''
\textit{Statistical Science},
vol.\ 17, no.\ 3, pp.\ 235--255, 2002.

\bibitem{pozzolo2015}
Dal Pozzolo, A., Caelen, O., Johnson, R.\ A., and Bontempi, G.,
``Calibrating Probability with Undersampling for Unbalanced Classification,''
in \textit{2015 IEEE Symposium Series on Computational Intelligence},
pp.\ 159--166, 2015.

\bibitem{sahin2011}
Sahin, Y.\ and Duman, E.,
``Detecting Credit Card Fraud by Decision Trees and Support Vector Machines,''
in \textit{Proceedings of the International MultiConference of Engineers and Computer Scientists (IMECS)},
vol.\ 1, pp.\ 1--6, 2011.

\end{thebibliography}
