\chapter*{Acknowledgement}
\addcontentsline{toc}{chapter}{Acknowledgement}
\thispagestyle{plain}

Working on this project has been one of the most challenging and, honestly, most rewarding things we have done during our four years here. There were moments when things did not work the way we expected, and there were also moments when everything clicked together perfectly. Through all of it, we had people around us who made the difference.

\vspace{0.2in}

\noindent Our deepest gratitude goes to our project supervisor, \textbf{Dr.\ Sushree Bibhuprada B.\ Priyadarshini}, Associate Professor, Department of CSIT, ITER, SOA University. She helped us understand not just the technical side of things, but also how to think about a research problem properly. Whenever we were stuck or going in the wrong direction, her guidance brought us back on track. Her patience with our questions and her willingness to explain things multiple times made a real difference to how this project turned out.

\vspace{0.2in}

\noindent We would also like to thank the \textbf{Head of the Department of Computer Science \& Information Technology} and all the faculty members who taught us over the past four years. The foundation we built in subjects like Machine Learning, Data Structures, Database Management Systems, and Computer Networks is what made this project possible.

\vspace{0.6in}

\noindent\textbf{Priyatosh Nayak}\\
\textbf{Smruti Ranjan Bhuyan}\\
\textbf{Krushnakanta Nayak}\\
\textbf{Bishmaya Bhuyan}
